# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r}
fname <- unzip("activity.zip")
df <- read.csv(fname, colClasses = c("integer", "Date", "integer"))
```

## What is mean total number of steps taken per day?
```{r}
steps_per_day <- aggregate(steps ~ date, data = df, FUN = sum, na.action = na.omit)

hist(steps_per_day$steps, 
        col = "red", 
        main = "Histogram of the total number of steps per day", 
        xlab = "Number of steps", 
        ylab = "Number of days", 
        breaks = 15)
```

```{r}
mean(steps_per_day$steps)
median(steps_per_day$steps)
```

### The mean from total number of steps taken per day is `r mean(steps_per_day$steps)`
### The median from total number of steps taken per day is `r median(steps_per_day$steps)`

## What is the average daily activity pattern?
```{r}
intervals_per_day <- aggregate(steps ~ interval, data = df, FUN = mean, na.action = na.omit)
plot(intervals_per_day,
        col = "red",
        xlab = "Interval",
        ylab = "Average number of steps taken",
        type = "l"
)
```

```{r}
max_steps <- intervals_per_day[with(intervals_per_day, which.max(steps)), "interval"]
max_steps
```

### Maximum number of steps represents the interval `r max_steps`.

## Imputing missing values
```{r}
missing <- subset(df, is.na(steps))
nrow(missing)
```

### The total number of missing values in the dataset is `r nrow(missing)`.

We fill the incomplete data with mean value per interval. This could be easily 
achieved by joining the dataset with *interval_per_day* variable:
```{r}
missing_without_steps <- missing[c("date", "interval")]
filled <- merge(missing_without_steps, intervals_per_day)
```

Merge to the complete data set
```{r}
complete <- rbind(filled, subset(df, !is.na(steps)))

steps_per_day <- aggregate(steps ~ date, data = complete, FUN = sum, na.action = na.omit)

hist(steps_per_day$steps, 
        col = "red", 
        main = "Histogram of the total number of steps taken each day", 
        xlab = "Number of steps", 
        ylab = "Number of days", 
        breaks = 15)
```

```{r}
mean(steps_per_day$steps)
median(steps_per_day$steps)
```

### The mean from total number of steps taken per day is `r mean(steps_per_day$steps)`
### The median from total number of steps taken per day is `r median(steps_per_day$steps)`
### As we can see, the filling NA values with mean for interval had only a slight impact causing almost no difference.

## Are there differences in activity patterns between weekdays and weekends?
A new factor variable in the dataset with two levels - "weekday" and "weekend" 
indicating whether a given date is a weekday or weekend day.

```{r}
weekinfo <- factor(
        ifelse(weekdays(complete$date) %in% c("Saturday", "Sunday"), 
               "weekend", "weekday"),
        levels = c("weekend", "weekday"))

complete <- cbind(complete, weekinfo)
```

Average number of steps taken, averaged across all days:
```{r}
library(lattice)
steps_per_day <- aggregate(steps ~ interval + weekinfo, data = complete, FUN = mean)
xyplot(steps ~ interval | weekinfo, data = steps_per_day, type = "l", layout = c(1, 2))
```
